---
title: "Appendix"
author: "Vicky Wang"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, include = FALSE}
# Load any packages you may need here
library(car)
library(emmeans)
```


```{r readData, include = FALSE}
# Save this template file in your final project's RStudio Project 
# Edit this code chunk to read in your data
fires <- read.csv("forestfires.csv")
fires$month <- as.factor(fires$month)
str(fires)
```

```{r fixyfixy, echo=FALSE}
fires <- na.omit(fires)
head(fires)
fires <- fires[, c(3,5:13)]
head(fires)
fires$month <- factor(fires$month, levels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))
```

***

#### **Figure 1: Numerical Summaries of Data**

```{r figure1, echo = FALSE}
summary(fires[2:10])
```

***

#### **Figure 2: Scatterplot Matrix of Data**

```{r figure2, echo=FALSE}
plot( ~ DMC + DC + ISI + temp + wind + area, data = fires,
      col = "darkorchid1")
```

***

#### **Figure 3: Histograms**

```{r figure3, echo=FALSE}
hist(fires$area,
     xlab = "Area Burned (in ha)",
     main = "Histogram of Burned Forest Area in Montesinho Park",
     col = "deeppink3",
     breaks = 50)

hist(log(fires$area),
     xlab = "log(Area Burned)",
     main = "Histogram of Log(Burn Area) in Montesinho Park",
     col = "deeppink4")
```

***

#### **Figure 4: Summary Output of Final Model**

```{r figure4, echo = FALSE}
lm_fires <- lm(log(area) ~ ISI + DC + DMC + wind + temp + month, data = fires)
summary(lm_fires)
```

***

#### **Figure 5: VIFs for Model 1**

```{r figure 5, echo = FALSE}
vif(lm_fires)
#multicollinearity secondary test
#summary(lm(ISI ~ DC + DMC + wind + temp, data = fires))$r.squared
#summary(lm(DC ~ ISI + DMC + wind + temp, data = fires))$r.squared
#summary(lm(DMC ~ DC + ISI + wind + temp, data = fires))$r.squared
#summary(lm(wind ~ DC + DMC + ISI + temp, data = fires))$r.squared
#summary(lm(temp ~ DC + DMC + wind + ISI, data = fires))$r.squared
```

***

```{r figure6, echo = FALSE}
# Include as many "figures" (model summaries, diagnostic plots, scatterplots by color, histograms, etc.) as necessary! But only include plots/output that you specifically reference in your report. 

# Each code chunk should have the argument `echo = FALSE` so that the code does not show up in the knitted file

# Each figure should be separated by three astersisks (***) 

# Each figure should be labeled using four hashtags (####) 

# If you have any questions, please don't hesitate to ask! This is worth a good chunk of points!

```

#### **Figure 6 Examining the Categorical Variable "Month"**

#### a) Does Month Affect Fire Size?
```{r plot, echo = FALSE}
plot(area ~ month, data = fires,
     xlab = "Month",
     ylab = "Burn Area Size",
     main = "Burn Area vs Month of Burn for Forest Fires",
     col = c("red", "orange", "yellow", "green", "chartreuse", "blue", "purple", "deeppink", "hotpink", "pink", "grey", "black"))

legend("topright",
       legend = levels(fires$month),
       col = c("red", "orange", "yellow", "green", "chartreuse", "blue", "purple", "deeppink", "hotpink", "pink", "grey", "black"),
       pch = 1,
       cex = 0.50)
```

#### b) Should we Include "Month" in the Model?
```{r compare monthinclusion, echo = FALSE}
lm_nomonth <- lm(log(area) ~ ISI + DC + DMC + wind + temp, data = fires)
summary(lm_nomonth)
lm_month <- lm(log(area) ~ ISI + DC + DMC + wind + temp + month, data = fires)
summary(lm_month)
```

#### **Figure 7 Month and Temperature Interaction**

#### a) Side-by-side of Temperature Ranges by Month
```{r plot2, echo = FALSE}
plot(fires$month, fires$temp,
     xlab = "Month",
     ylab = "Temperature (C)",
     main = "Months and Temperatures for Fires",
     col = c("coral", "coral1", "coral2", "coral3", "coral4", "brown", "brown2", "brown3", "brown", "brown4"))

legend("topright",
       legend = levels(fires$month),
       col = c("coral", "coral1", "coral2", "coral3", "coral4", "brown", "brown2", "brown3", "brown", "brown4"),
       pch = 1,
       cex = 0.50)

```
#### b) Comparison of Interaction Terms with Month
```{r monthinclusions, echo = FALSE}
lm_nointeraction <- lm(log(area) ~ ISI + DC + DMC + wind + temp + month, data = fires)
lm_interaction <- lm(log(area) ~ ISI + DC + DMC + wind + temp*month, data = fires)

summary(lm_nointeraction)
summary(lm_interaction)

AIC(lm_nointeraction)
AIC(lm_interaction)

```

#### c) Scatterplot of between Area and Temperature by Month
```{r colorbymonth, echo = FALSE}
plot(area ~ temp,
     data = fires,
     col = month,
     xlab = "Temp (c)",
     ylab = "Area (ha)",
     main = "Scatterplot of Area vs Temperature by Month")
```



#### **Figure 7 OLS Assumptions**

#### a) Residuals vs Fitted Values for lm_fires
```{r lm_fires, echo = FALSE}
plot(lm_fires$fitted.values, lm_fires$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values Plot",
     col = "darkslategray4")
abline(h = 0, lty = 2, col = "deeppink")
```


#### b) QQ Plot for lm_fires
```{r qqplot, echo = FALSE}
qqnorm(lm_fires$residuals, main = "QQ Plot of the Residuals", col = "deeppink")
qqline(lm_fires$residuals, col = "purple")
```
